\chapter{Quantum Monte Carlo}

\section{Modeling Diffusion}

Like any phenomena involving a density function, or distribution, Quantum Mechanics can be modeled by diffusion. In Quantum Mechanics, the distribution is given by $|\p|^2$, the Wave function squared. The diffusing elements of interest are the particles making up our system. The idea is to have an ensemble of \textit{Random Walkers} in which each walker represents a position is space (and time for time-dependent studies). Averaging values over the paths of the ensemble will yield average values corresponding to the probability distribution governing the movement of individual walkers. 

Such random movement is referred to as a \textit{Brownian motion}, named after the British Botanist R. Brown, originating from his experiments on plant pollen dispersed in water. \textit{Markov chains} are a subtype of Brownian motion, where a walkers next move is independent of previous moves. This is the stochastic process in which Quantum Monte Carlo is described.

The purpose of this section is to motivate the use of diffusion theory in Quantum Mechanics, and to derive the sampling rules needed in order to model Quantum Mechanical distributions by diffusion of random walkers correctly. I will be using natural units, that is $\hbar$, $m_e$, etc. are all set to unity, in order to simplify the expressions.

\subsection{Stating the \schrodinger as a Diffusion Problem}

Consider the time-dependent \schrodinger for an abstract state $\phi(x, t)$ with an arbitrary energy shift $E'$

\begin{equation}
 -\frac{\partial \phi(x, t)}{i\partial t} = (\OP{H} - E')\phi(x, t).
\end{equation}

The formal solution given a time-independent Hamiltonian is given by separation of variables in $\phi(x, t)$ (see \cite{griffiths} for more details regarding assumptions etc.). This corresponds to using the standard time-evolution operator. Further expanding $\phi(x, t=t_0)$ in the eigenstates of $\OP{H}$, $\Phi_k(x)$, yields the following solution

\begin{eqnarray}
 \phi(x, t) &=& e^{-i(\OP{H} - E')(t-t_0)}\phi(x, t=t_0) \label{eq:schrodGeneralSolution} \\
            &=& \sum_{k=0}^{\infty} C_k\Phi_k(x)e^{-i(E_k - E')(t-t_0)} \label{eq:realSchrod} \\
 C_k &=& \braket{\Phi_k(x)}{\phi(x, t=0)} \nonumber
 \end{eqnarray}


The real form of this equation is obtained through a Wick rotation in time ($it \rightarrow \tau$), which basically means that the time-evolution operation is substituted by a \textit{projection operation}. To illustrate this, look at the solution of the real equation with $E'$ chosen to be the ground state energy of $\OP{H}$, and $t_0$ chosen to be zero

\begin{eqnarray}
 \phi(x, \tau) &=& \sum_{k=0}^{\infty} C_k\Phi_k(x)e^{-(E_k - E_0)\tau} \label{eq:schrodGeneralSolution3}\\
                    &=& C_0\Phi_0(x) + \sum_{k=1}^{\infty} C_k\Phi_k(x)e^{-\delta E_k\tau}, \label{eq:schrodGeneralSolution2}
\end{eqnarray}

where $\delta E_k = E_k - E_0 > 0$ for $k \ge 1$. Observe that as $\tau$ increase, the excited states of $\OP{H}$ will be exponentially dampened, hence the name projection operation. Note however, that no real-time solutions can be achieved by solving this equation; only in the limit $\tau\rightarrow\infty$ will the solution match that of the \schrodinger. 

The approach of Quantum Monte Carlo is to split the evolution of $\tau$ into small sequential steps $\delta\tau$, and use the latest calculated energy, the \textit{trial energy}, $E_T$, as an approximation to the ground state energy. The error introduced by this approximation will be discussed in Section~\ref{sec:DMC}. Either way, restating Eq.~(\ref{eq:schrodGeneralSolution3}) in terms of sequential steps yields

\begin{equation}
  \phi(x, n\delta\tau) = \sum_{k=0}^{\infty} C_k\Phi_k(x)e^{-(E_k - E_T)n\delta\tau} \label{eq:genSolTimeSteps},
\end{equation}

which remains exact for an infinite amount of infinitely small time steps. In practice, a balance is sought between the \textit{number of cycles}, $n$, and the time step. This restatement might seem redundant, however, it is the key which opens up the possibility to model the solution by discretized diffusion of particles in Quantum Mechanical potentials. 

\subsection{Solving the Diffusion Problem}

In this chapter \textit{Dirac Notation} will be used. See Appendix \ref{app:Dirac} for an introduction.

Consider two sequential time steps $\tau$ and $\tau + \delta\tau$ in Eq.~(\ref{eq:genSolTimeSteps}). Without expanding the current state $\ket{\phi(\tau)}$ like previously, we are left with an expression on the form

\begin{equation*}
 \ket{\phi(\tau + \delta\tau)} = e^{-(\OP{H} - E_0)\delta\tau}\ket{\phi(\tau)}.
\end{equation*}

Taking the inner product with $\bra{x}$ on both sides and inserting a complete set of states $\ket{y}$ on the left hand side yields

\begin{eqnarray*}
 \phi(x, \tau + \delta\tau) &=& \int \bra{x}e^{-(\OP{H} - E_0)\delta\tau}\ket{y}\phi(y, \tau)dy \\
			     &\equiv& \int G(x, y; \delta\tau)\phi(y, \tau)dy
\end{eqnarray*}

where the \textit{Green's Function}, $G(x, y; \delta\tau)$ serves as a transition probability. This is the path integral formalism of Quantum Mechanics, where a final state is obtained by integrating through all possible paths of all the particles in the system, but I will not go into details about it here. More information about this can be found in \cite{leinaas}. A more detailed derivation of the Green's Function are given in \cite{abInitioMC}.

This is still not a practical way of solving the equations. In finding the Green's function we might as well find the solution directly. However, what we can do, is make the time steps very small and split the Hamiltonian into the kinetic, $\OP{T}$ and the potential part, $\OP{V}$:

\begin{eqnarray}
 \bra{x}e^{-(\OP{H} - E_0)\delta\tau}\ket{y} &=& \bra{x}e^{-(\OP{T} + \OP{V} - E_T)\delta\tau}\ket{y} \nonumber \\
                                             &=& \bra{x}e^{-\OP{T}\delta\tau}e^{-(\OP{V} - E_T)\delta\tau}\ket{y} + \frac{1}{2}[\OP{V}, \OP{T}]\delta\tau^2 + \mathcal{O}(\delta\tau^3) \nonumber \\
                                             &\simeq& \bra{x}e^{-\OP{T}\delta\tau}e^{-(\OP{V} - E_T)\delta\tau}\ket{y} 
\end{eqnarray}

The last step is known as the \textit{short time approximation}, that is, we split the Green's Function into two processes which are well known. The first which represents a diffusion process

\begin{equation}
\label{eq:GDiff}
 G_\mathrm{Diff} = e^{-\OP{T}\delta\tau}
\end{equation}

and the second which represents a branching process

\begin{equation}
 \label{eq:GB}
 G_\mathrm{B} = e^{-(\OP{V} - E_T)\delta\tau}
\end{equation}

Once a diffusion model is selected, for each cycle, the Green's functions are used to propagate the ensemble of walkers making up our distribution into the next time step. The final distributions of walkers will correspond to that of the direct solution of the \schrodinger, given that the time step is sufficiently small, and we simulate long enough. These constraints will be covered in more detail later. 

Incorporating only the effect of Eq.(\ref{eq:GDiff}) results in a method called \textit{Variational Monte Carlo}. Including the branching term as well results in \textit{Diffusion Monte Carlo}. These methods will be discussed in sections \ref{sec:VMC} and \ref{sec:DMC}. In either of these methods, diffusion is the key process. In practice, we choose a diffusion model where closed form expressions for the Green's Functions exists. Two examples of this will be presented in the following sections.



\subsection{Isotropic Diffusion}

Isotropic diffusion is a process in which diffusing particles sees all possible directions as an equally probable path. Eq.~(\ref{Eq:diffusionSimple}) is an example of this. This is the simplest form of a diffusion equation, the case with a linear \textit{diffusion constant}, $D$, and no drift terms.

\begin{equation}
 \label{Eq:diffusionSimple}
 \frac{\partial P(\vec r, t)}{\partial t} = D\nabla^2 P(\vec r, t) 
\end{equation}

In the Quantum Monte Carlo case, the value of the diffusion constant is $D=\frac{1}{2}$, that is, the term scaling the Laplacian in the \schrodinger. The closed form expression for the Green's function is a Gaussian distribution with variance $2D\delta t$ \cite{abInitioMC}

\begin{equation}
\label{eq:GF_iso}
 G_{\mathrm{Diff}}^{\mathrm{ISO}}(i\,\rightarrow j) \propto e^{-(x_i-x_j)^2/4D\delta\tau}  .
\end{equation}

These equations describe the diffusion process theoretically, however, in order to achieve specific sampling rules for our walkers, we need a connection between the time-dependence of the total distribution and the time-dependence of an individual walker's components in configuration space. This connection is given in terms of a stochastic differential equation called \textit{The Langevin Equation}.

\subsubsection{The Langevin Equation for Isotropic Diffusion}

The Langevin Equation is a stochastic differential equation used in physics to relate the time dependence of a distribution to the time-dependence of the degrees of freedom in the system. For the simple isotropic diffusion described previously, solving the Langevin equation using a Forward Euler approximation for the time derivative results in the following relation:

\begin{eqnarray}
\label{eq:langevinSolSimple}
 x_{i+1} = x_i + \xi, \qquad\qquad \mathrm{Var}(\xi) &=& 2D\delta t, \\
			     \langle\xi\rangle &=& x_i, \nonumber 
\end{eqnarray}

where $\xi$ is a normal distributed number whose variance match that of the Green's function in Eq.~(\ref{eq:GF_iso}). This relation is in agreement with the isotropy of Eq.~(\ref{Eq:diffusionSimple}) in the sense that the displacement is symmetric around the current position.


\subsection{Anisotropic Diffusion: Fokker-Planck}

Anisotropic diffusion, in contrast to isotropic diffusion, does not count all directions as equally probable. An example of this is diffusion according to the \textit{Fokker-Planck Equation}, that is, diffusion with a drift term, $\vec F(\vec r, t)$, responsible for pushing the walkers in the direction of configurations with higher probabilities.

\begin{equation}
 \label{Eq:fokkerPlanck}
 \frac{\partial P(\vec r, t)}{\partial t} = D\nabla\cdot\Big[\Big(\nabla - \vec F(\vec r, t)\Big) P(\vec r, t)\Big] 
\end{equation}

The remarkable thing is that simple isotropic diffusion processes obey this relation \cite{abInitioMC}. This means that Quantum Mechanical distributions can be modeled by the Fokker-Planck Equation, leading to a more optimized way of sampling in practical situations. This method of \textit{Importance Sampling} will be discussed in Section~(ref imp). 

In Quantum Monte Carlo we want convergence to a stationary state. We can use this criteria to deduce expression for the drift term given our Quantum Mechanical distribution. A stationary state is obtained when the left hand side of Eq.~(\ref{Eq:fokkerPlanck}) is zero:

\begin{equation*}
 \nabla^2 P(\vec r, t) = P(\vec r, t)\nabla\cdot\vec F(\vec r, t) + \vec F(\vec r, t) \cdot \nabla P(\vec r, t)
\end{equation*}

The next thing we want to achieve is cancellation in the rest of the terms. In order to obtain a Laplacian term on the right hand side to potentially cancel out the one on the left, the drift term needs to be on the form $F(\vec r, t) = g(\vec r, t)\nabla P(\vec r, t)$. Inserting this yields

\begin{equation*}
  \nabla^2 P(\vec r, t) = P(\vec r, t)\frac{\partial g(\vec r, t)}{\partial P(\vec r, t)}\Big|\nabla P(\vec r, t)\Big|^2
  + P(\vec r, t)g(\vec r, t)\nabla^2 P(\vec r, t) + g(\vec r, t) \Big|\nabla P(\vec r, t)\Big|^2.
\end{equation*}

Looking at the factors in front of the Laplacian suggests using $g(\vec r, t) = 1/P(\vec r, t)$. A quick check reveals that this also cancels out the gradient terms, and the resulting expression for the drift term becomes

\begin{eqnarray}
 \vec F(\vec r, t) &=& \frac{1}{P(\vec r, t)}\nabla P(\vec r, t) \nonumber \\
                   &=& \frac{2}{|\psi(\vec r, t)|}\nabla |\psi(\vec r, t)|
\end{eqnarray}

In Quantum Monte Carlo, the drift term is called \textit{The Quantum Force}, since it is responsible for pushing the walkers into regions of higher probabilities, analogous to a force in Newtonian mechanics.

Another strength of the Fokker-Planck equation is that even though the equation itself is more complicated, it's Green's function still has a closed form solution. This means that we can evaluate it efficiently. If this was not the case, the practical value would be reduced dramatically. The reason for this will become clear in Section \ref{sec:MetroMain}. As expected, it is no longer symmetric 

\begin{equation}
\label{eq:GF_FP}
 G_\mathrm{Diff}^\mathrm{FP}(i\,\rightarrow j) \propto e^{-(x_i-x_j - D\delta\tau F(x_i))^2/4D\delta\tau}.
\end{equation}


\subsubsection{The Langevin Equation for the Fokker-Planck Equation}

The Langevin equation in the case of a Fokker-Planck Equation has the following form

\begin{equation}
 \frac{\partial x_i}{\partial t} = D F(x_i) + \eta,
\end{equation}

where $\eta$ is a so-called \textit{noise term} from stochastic processes. Solving this using the same method as for the isotropic case yields the following sampling rules

\begin{equation}
 \label{eq:langevinSolFP}
 x_{i+1} = x_i + \xi + DF(x_i)\delta t,
\end{equation}

where $\xi$ is the same as for the isotropic case. We observe that if the drift term is set to zero, we are back in the isotropic case, just as required. For more details regarding the Fokker-Planck Equation and the Langevin equation, see \cite{Gardiner:2004bk}, \cite{risken1989fpe} and \cite{langevin}.


\section{Diffusive Equilibrium Constraints}

Upon convergence of a Markov process, the system at hand will reach its most likely state. This is exactly the behaviour of a real system of diffusing particles described by statistical mechanics: It will \textit{thermalize}, that is, be in its most likely state given a surrounding temperature. Markov processes are hence beloved by physicists; they are simple, yet realistic. 

Once thermalization is reached, average values may be sampled. However, simply spawning a Markov process and waiting for thermalization is an inefficient and unpractical scenario. This may take forever, and it may not; either way its not optimal. We can introduce rules of acceptance and rejection of transitions purposed by following the solutions of the Langevin Equation (Eq.~(\ref{eq:langevinSolSimple}) and Eq.~(\ref{eq:langevinSolFP})), but not without constraints. If any of the following conditions break, we have no guarantee that the system will thermalize properly:

\subsection{Detailed Balance} 

For Markov processes, detailed balance is achieved by demanding a \textit{Reversible} Markov process. This boils down to a statistical requirement stating that 

\begin{equation}
 \label{eq:DetailedBalance}
 P_iW(i\,\rightarrow\,j) = P_jW(j\,\rightarrow\,i),
\end{equation}

where $P_i$ is the probability density in configuration $i$, and $W(i\,\rightarrow\,j)$ is the transition probability between states $i$ and $j$. 

\subsection{Ergodicity}

Another requirement is that the sampling must be ergodic, that is, the Markov chain (or simpler: The walker) needs to be able to be able to reach any configuration state in the space spanned by the distribution function. It is tempting to define a brute force acceptance rule where only steps resulting in a higher overall probability is accepted, however, this limits the path of the walker, and hence breaks the ergodicity requirement.

\section{The Metropolis Algorithm}
\label{sec:MetroMain}

The Metropolis Algorithm is a simple set of acceptance/rejection rules used in order to make the thermalization more effective. I will not go into details about transport theory in this section, but rather start the derivation from the criteria of detailed balance, Eq.~(\ref{eq:DetailedBalance}), modeling the transition probability as two-part: $g(i\,\rightarrow\,j)$, the probability of selecting configuration $j$ given configuration $i$, times a probability of accepting the selected move, $A(i\,\rightarrow\,j)$:

\begin{eqnarray}
 \label{eq:metro1}
 P_iW(i\,\rightarrow\,j) &=& P_jW(j\,\rightarrow\,i) \nonumber \\
 P_ig(i\,\rightarrow\,j)A(i\,\rightarrow\,j) &=& P_jg(j\,\rightarrow\,i)A(j\,\rightarrow\,i)
\end{eqnarray}

Inserting the probability distribution as the Wave function squared, and the selection probability as our Green's function, then gives us a simple expression:

\begin{eqnarray}
  \label{eq:metro2}
  |\psi_i|^2G(i\,\rightarrow\,j)A(i\,\rightarrow\,j) &=& |\psi_j|^2G(j\,\rightarrow\,i)A(j\,\rightarrow\,i) \nonumber \\
  \frac{A(j\,\rightarrow\,i)}{A(i\,\rightarrow\,j)} &=& \frac{G(i\,\rightarrow j)}{G(j\,\rightarrow i)}\frac{|\psi_i|^2}{|\psi_j|^2} \equiv R_G(j\,\rightarrow\,i)R_\psi(j\,\rightarrow\,i)^2
\end{eqnarray}

Assume now that configuration $i$ has a higher overall probability than configuration $j$, that is, $A(i\,\rightarrow\,j) = 1$. A more effective thermalization is obtained by accepting all these moves. What saves us from breaking the criteria of ergodicity is the fact that we do not reject the move otherwise. Instead, we insert $A(i\,\rightarrow\,j) = 1$ into Eq.~(\ref{eq:metro2}), and thus ensuring detailed balance as well as ergodicity. This yields

\begin{equation*}
 A(j\,\rightarrow\,i) = R_G(j\,\rightarrow\,i)R_\psi(j\,\rightarrow\,i)^2
\end{equation*}


Concatenating both scenarios yields the following acceptance/rejection rules:

\begin{equation}
\label{eq:MetroGeneralGreen}
 A(i\,\rightarrow\,j) = \left\{\begin{array}{ccc}
R_G(i\,\rightarrow\,j)R_\psi(i\,\rightarrow\,j)^2 & & R_G(i\,\rightarrow\,j)R_\psi(i\,\rightarrow\,j)^2 < 1\\
1  & & \mathrm{else}  \end{array}\right. 
\end{equation}

Or more simplistic:

\begin{equation}
  A(i\,\rightarrow\,j) = \min\{R_G(i\,\rightarrow\,j)R_\psi(i\,\rightarrow\,j)^2, \,1\}
\end{equation}


For the isotropic case, we have a cancellation of the Greens function due to symmetry, $R_G(i\,\rightarrow\,j) = 1$, leaving us with the standard Metropolis algorithm:

\begin{equation}
\label{eq:Metropolis_standard}
 A(i\,\rightarrow\,j) = \min\{R_\psi(i\,\rightarrow\,j)^2, \,1\}
\end{equation}

If we on the other hand use the Fokker-Planck equation, we will not get a cancellation. Inserting Eq.~(\ref{eq:GF_FP}) into Eq.~(\ref{eq:MetroGeneralGreen}) results in the \textit{Metropolis Hastings algorithm}. The ratio of Green's function can be evaluated efficiently by simply subtracting the exponents of the exponentials

\begin{equation}
 R_G^\mathrm{FP}(i\,\rightarrow\,j) = G_\mathrm{Diff}^\mathrm{FP}(j\,\rightarrow i)/G_\mathrm{Diff}^\mathrm{FP}(i\,\rightarrow j) = e^{\frac{1}{2}\big(F(x_j) + F(x_i)\big)\big(\frac{1}{2}D\delta t(F(x_j) - F(x_i)) + x_i - x_i\big)}
\end{equation}

\begin{equation}
\label{eq:MetropolisHastings}
 A(i\,\rightarrow\,j) = \min\{R_G^\mathrm{FP}(i\,\rightarrow\,j)R_\psi(i\,\rightarrow\,j)^2, \,1\}
\end{equation}

Derived from detailed balance, the Metroplis Algorithm is as a must-have when it comes to Markov Chain Monte Carlo. Besides Quantum Monte Carlo, we have methods such as the \textit{Ising Model} which greatly benefit from these rules \cite{morten}.

\section{The Trial Wave Function}

Recall Eq.~(\ref{eq:schrodGeneralSolution})-(\ref{eq:schrodGeneralSolution2}). Before we start propagating the distribution  in time, we need the boundary condition $\phi(\vec r, t=0)$, from here on referred to as the trial wave function. Mathematically, we may choose any normalizable wave function, whose overlap with the exact wave function, $\Phi(x)$, is non-zero. If the overlap is zero, that is, $C_0=0$ in Eq.~(\ref{eq:schrodGeneralSolution2}), the entire diffusion formalism breaks down, and no final state of convergence can be reached. The opposite is also the case; the closer $C_0$ is to unity, the more rapidly it will become the dominant contribution to our distribution. A clever choice of Trial Wave function will greatly improve the rate of convergence.

\subsection{Choice of Trial Wave function}

Picture having a superposition of states making up the trial wave function, tweaked and tuned so that the overlap is extremely close to unity. At this point, a lot of computational time may have been spent already in preparation of the   which converges within a given criteria in $N$ cycles. On the other hand, with a simplistic wave function consisting of a single state, the same criteria is reached in $5N$ cycles, however, each cycle taking only a fraction of the original CPU-time; it's a balance. 

% It should be mentioned this is all depending on the system at hand. For interacting electrons in a harmonic oscillator (Quantum Dots), a single Slater determinant consisting of standard harmonic oscillator eigenstates is sufficient. The same works for Atoms and hydrogen-like wave functions. However, if we were to model a nucleus, no choice of basis is more obvious than another. In this case one has to introduce superpositions of slater determinants constructed from e.g. a free particle basis in order to obtain reasonable convergence.

The simplistic wave function is an ideal choice when:

\begin{itemize}
 \item The system closely resembles that of another analytically solvable problem, e.g. Harmonic Oscillator or Hydrogen.
 \item The number of degrees of freedom is large. Simplistic wave functions can be optimized to scale extremely well with system size.
\end{itemize}

If we on the other hand are outside these criteria, a more complicated trial wave function should be considered.

SKRIVESPERREEEE




\section{Variational Monte-Carlo}
\label{sec:VMC}



\section{Diffusion Monte-Carlo}
\label{sec:DMC}




