\chapter*{Preface}

During my stay at junior high school, I had no personal interest in mathematics, scoring average grades and skipping homework. When the final year was finished, I sought an education within the first love of my life: Music. However, sadly, my application was turned down, forcing me to fall back on general topics. This is where I had my first encounter with physics, which should turn out not to be the last.

When high school ended, I applied for an education within structural engineering at the Norwegian University of Science and Technology, however, again I was turned down. Second on the list was the physics, meteorology and astronomy program at the University of Oslo. Never in my life have I been more grateful for being turned down, as starting at the University of Oslo introduced me to the second love of my life: Programming. 

During the third year of my Bachelor I had a hard time figuring out which courses I should pick. Quite randomly, I attended a course on computational physics lectured by Morten Hjorth-Jensen. It immediately hit me that he had a genuine interest in the well-being of his students. It did not take long before I decided to apply for a master within the field of computational physics.

To summarize, me being here today writing this thesis is a result of extremely random events. However, I could not have landed in a better place, and I am forever grateful for my time here at the University of Oslo. So grateful in fact, that I am continuing my stay as a PhD student within the field of \textit{multi-scale physics}.

I would like to thank my fellow Master students Sarah Reimann, Sigve B\o{}e Skattum, and Karl Leikanger for much appreciated help and good discussions. Sigve helped me with the minimization algorithm, that is, the darkest chapter of this thesis. A huge thanks to Karl for all those late-nights where you gave me a ride home. Finally, thank you Sarah for withstanding my horrible German phrases all these years. 

My supervisor, Morten Hjorth-Jensen, deserves a special thanks for believing in me, and filling me with confidence when it was much needed. This man is awesome.  

Additionally I would like to thank the rest of the people at the computational physics research group, especially Mathilde N. Kamperud and Veronica K. B. Olsen whom with I shared my office, Svenn-Arne Dragly for helping me with numerous Linux-related issues, and ``brochan'' Milad H. Mobarhan for all the pleasant hours spent at Nydalen Kebab. You all contribute to making this place the best it can possibly be. Thank you!


\hfill J\o{}rgen H\o{}gberget

Oslo, June 2013